---
title: "RNA-seq analysis of mga2 and cbf11 mga2 mutants in YES and YES+ammonium"
author: "Martin Převorovský, prevorov@natur.cuni.cz, http://www.GenoMik.cz"
output: html_document
---

```{r setup, message=FALSE}
library(rtracklayer)
library(GenomicAlignments)
library(DESeq2)
library(RColorBrewer)
library(pheatmap)

knitr::opts_chunk$set(echo = TRUE)

genome_dir <- './genome/'
genome_gff <- paste0(genome_dir, 'Schizosaccharomyces_pombe_all_chromosomes.gff3')
bam_dir <- './BAM_RNA-seq/'
coverage_dir <- './coverage_RNA-seq/'
image_dir <- './images/'
deg_dir <- './DEG_RNA-seq/'
```

Variables for shell scripts are stored in the file "shell_variables_RNA-seq.sh" and can be modified as needed:

```{bash list_variables}
cat shell_variables_RNA-seq.sh
```


# Getting the raw RNA-seq read datasets

THIS SECTION WILL ONLY BECOME RELEVANT ONCE THE SEQUENCING DATA ARE PUBLIC in ARRAYEXPRESS

All data and metadata can be obtained from ArrayExpress at https://www.ebi.ac.uk/arrayexpress/experiments/XXXXXXXX/.

```{bash, eval=FALSE, message=FALSE}
source shell_variables_RNA-seq.sh

wget -N --no-verbose https://www.ebi.ac.uk/arrayexpress/files/XXXXXXXX/XXXXXXXX.sdrf.txt
mapfile -t ftp_links < <(tail -n +2 XXXXXXXX.sdrf.txt | cut -f30)
mapfile -t ftp_names < <(tail -n +2 XXXXXXXX.sdrf.txt | cut -f28)

mkdir -p "${fastq_dir}"
for i in ${!ftp_links[@]};
do
  wget --no-verbose -O "${fastq_dir}${ftp_names[${i}]}" "${ftp_links[${i}]}"
done
```


# *Schizosaccharomyces pombe* genome sequence and annotation

* genome was downloaded from ftp://ftp.pombase.org/pombe/genome_sequence_and_features/genome_sequence/ (downloaded `r date()`)

* genome annotation (GFF): ftp://ftp.pombase.org/pombe/genome_sequence_and_features/gff3/ (downloaded `r date()`)

```{bash}
bedtools --version
```

```{bash, message=FALSE, eval=TRUE}
source shell_variables_RNA-seq.sh

mkdir -p "${genome_dir}"
wget -N --no-verbose --directory-prefix="${genome_dir}"  ftp://ftp.pombase.org/pombe/genome_sequence_and_features/genome_sequence/Schizosaccharomyces_pombe_chromosome_I.fa.gz
wget -N --no-verbose --directory-prefix="${genome_dir}"  ftp://ftp.pombase.org/pombe/genome_sequence_and_features/genome_sequence/Schizosaccharomyces_pombe_chromosome_II.fa.gz
wget -N --no-verbose --directory-prefix="${genome_dir}"  ftp://ftp.pombase.org/pombe/genome_sequence_and_features/genome_sequence/Schizosaccharomyces_pombe_chromosome_III.fa.gz
wget -N --no-verbose --directory-prefix="${genome_dir}" ftp://ftp.pombase.org/pombe/genome_sequence_and_features/genome_sequence/Schizosaccharomyces_pombe_mitochondrial_chromosome.fa.gz
zcat `ls "${genome_dir}"* | grep fa.gz$` > "${genome}"

wget -N --no-verbose --directory-prefix="${genome_dir}"  ftp://ftp.pombase.org/pombe/genome_sequence_and_features/gff3/Schizosaccharomyces_pombe_all_chromosomes.gff3.gz
gzip -fd "${genome_dir}Schizosaccharomyces_pombe_all_chromosomes.gff3.gz"

samtools faidx "${genome}"
cut -f1,2 "${genome}.fai" > "${genome_dir}chrom.sizes"
```


# QC of raw sequence reads

Technical quality of library construction and sequencing run was tested using fastQC

```{bash}
fastqc --version
multiqc --version
```

```{bash, eval=TRUE}
source shell_variables_RNA-seq.sh

mkdir -p "${QC_dir}"
mapfile -t fastq_files < <(ls -1 -d "${fastq_dir}"* | grep "${fastq_file_ext}")
fastqc --quiet --outdir "${QC_dir}" --threads "${CPU}" "${fastq_files[@]}"
multiqc -f -m fastqc -i RNA-seq_mga2 -o "${QC_dir}" "${QC_dir}"
```

Results of QC:

* 76 nt SE reads

* 23.8-46.8M reads per sample

* adapter contamination below ~1% for most samples, 1 sample has ~5%. Will be trimmed.

* all samples give error on high read duplication levels (normal and OK for RNA-seq from a small genome)

* all samples give warning on per base sequence content (positions 1-13) (normal for random-primed cDNA)

* 4 samples give warning on GC content, likely due to polyA reads

* 10 samples gives warning/error on overrepresented sequences (polyA reads, adapters)

* read quality scores deteriorate towards the 3' end. Trimming is needed.


# Read trimming

* remove Illumina adapters

* trim parts of the reads where average base quality in a 4 nt window drops below 20

* only keep trimmed reads that are at least 20 nt long

```{bash}
java -jar /opt/Trimmomatic-0.39/trimmomatic-0.39.jar -version
```

```{bash, eval=TRUE}
source shell_variables_RNA-seq.sh

mkdir -p "${fastq_trim_dir}"
mapfile -t fastq_files < <(ls -1 "${fastq_dir}" | grep "${fastq_file_ext}")
for i in ${fastq_files[@]};
do
  java -jar /opt/Trimmomatic-0.39/trimmomatic-0.39.jar SE -phred33 -threads "${CPU}" "${fastq_dir}${i}" "${fastq_trim_dir}${i}.trim.fastq.gz" ILLUMINACLIP:/opt/Trimmomatic-0.39/adapters/TruSeq3-PE-2.fa:2:30:10:2 SLIDINGWINDOW:4:20 MINLEN:20
done
```


# QC of quality-trimmed sequence reads

Effects of trimming were determined by `fastQC`.

```{bash, eval=TRUE}
source shell_variables_RNA-seq.sh

mkdir -p "${QC_trim_dir}"
mapfile -t fastq_files < <(ls -1 -d "${fastq_trim_dir}"* | grep "${fastq_file_ext}")
fastqc --quiet --outdir "${QC_trim_dir}" --threads "${CPU}" "${fastq_files[@]}"
multiqc -f -m fastqc -i RNA-seq_mga2_trimmed -o "${QC_trim_dir}" "${QC_trim_dir}"
```

Results of QC:

* dropped ~2-4.5% of reads (typically 2-3%)

* 20-76 nt reads (vast majority of reads are >72 nt)

* read quality scores now OK

* adapters were removed


# Read mapping into *S. pombe* genome

```{bash}
hisat2 --version
samtools --version
```

```{bash, eval=TRUE}
source shell_variables_RNA-seq.sh

# build HISAT2 index
hisat2-build --quiet -p "${CPU}" "${genome}" "${genome}"

# map
mkdir -p "${bam_dir}"
mapfile -t fastq_files < <(ls -1 "${fastq_trim_dir}" | grep "${fastq_file_ext}")
for i in ${fastq_files[@]};
do
	infile="${fastq_trim_dir}${i}"
	outfile="${bam_dir}${i}.bam"
  echo "${outfile}"
	hisat2 -x "${genome}" -U "${infile}" --threads "${CPU}" --max-intronlen 10000 -k 15 --summary-file "${outfile}.log" | samtools view -b -F 256 --threads "${CPU}" - | samtools sort --threads "${CPU}" -o "${outfile}" - 
	samtools index "${outfile}"
	samtools view -c --threads "${CPU}" "${outfile}"
done
```

Notes

* since signal is expected at repeat regions, HISAT was set to search for up to 15 matching positions (this allows even for testing all Tf2 and dh/dg repeats in the genome), then only the best (marked as primary by HISAT) alignment was kept

* **The kit used for sequencing library preparation produces reads that are reverse-complement to the original RNA sequence. Therefore, reads map to the opposite strand compared to the corresponding gene's orientation. This needs to be accounted for during count table generation (reads are reversed) and generation of strand-specific BAM and BigWig files (plus-strand reads are used for computing real minus-strand coverage, and vice versa).**
	
* Mapping rate >97%.


# Calculate genome coverage tracks

```{bash, eval=TRUE}
source shell_variables_RNA-seq.sh

mkdir -p "${coverage_dir}"
mapfile -t bam_files < <(ls -1 "${bam_dir}" | grep "\.bam$")
for i in ${!bam_files[@]};
do
	input_file="${bam_dir}${bam_files[${i}]}"
	bamCoverage --binSize "${bin_size}" --normalizeUsing CPM -p "${CPU}" --bam "${input_file}" -o "${coverage_dir}${bam_files[${i}]}.bw"
done

# average coverage tracks across biological replicates
mapfile -t bw_files < <(ls -1 "${coverage_dir}" | grep "\.bam\.bw$" | sort -V)
samples=("mga2_YES" "mga2_YES+AMM" "mga2_cbf11_YES" "mga2_cbf11_YES+AMM")
for i in {0..3};
do
  wiggletools mean "${coverage_dir}${bw_files[${i}]}" "${coverage_dir}${bw_files[${i}+4]}" "${coverage_dir}${bw_files[${i}+8]}" | wigToBigWig stdin "${genome_dir}chrom.sizes" "${coverage_dir}${samples[${i}]}.mean.bw"
done
```

Notes

* Checked the genotypes visually in IGV → OK.


# Identify differentially expressed genes

## Generate gene-level counts tables.

* Raw count data are needed for DEG calling in `DESeq2`.

* TPM-normalized (Transcripts Per Million) count data are suitable for general comparisons of the samples.

```{r, eval=TRUE}
gff <- import.gff3(genome_gff)
gff.genes <- gff[gff$type == 'gene', ]

sample_info <- read.delim('sampleInfo', stringsAsFactors = FALSE, header = TRUE, sep = '\t')
sample_info[, 'cultivation_run'] <- factor(sample_info[, 'cultivation_run'])
sample_info[, 'genotype'] <- factor(sample_info[, 'genotype'])
sample_info[, 'medium'] <- factor(sample_info[, 'medium'])

bam <- BamFileList(list.files(bam_dir, 
                              pattern = '\\.bam$', 
                              full.names = TRUE),
                   asMates = FALSE,
                   yieldSize = 1000000)
names(bam) <- sample_info$sample_ID

se <- summarizeOverlaps(gff.genes,
                        bam,
                        mode = 'Union',
                        singleEnd = TRUE,
                        ignore.strand = FALSE,
                        preprocess.reads = invertStrand)
rownames(se) <- gff.genes$ID
colData(se) <- cbind(colData(se), sample_info)
#save(se, file = 'se.rda')
#load(file = 'se.rda')

counts <- assay(se)

# normalize to get TPM values (transcripts per million)
gene_widths <- width(gff.genes)
counts.norm <- counts
for (i in 1:nrow(counts.norm)){
  counts.norm[i, ] <- counts.norm[i, ] / gene_widths[i]
}
count_sums <- colSums(counts.norm) / 1000000
for (i in 1:length(count_sums)){
  counts.norm[, i] <- counts.norm[, i] / count_sums[i]
}

counts <- cbind(gff.genes$Name, counts)
counts.norm <- cbind(gff.genes$Name, counts.norm)
colnames(counts)[1] <- 'gene_name'
colnames(counts.norm)[1] <- 'gene_name'
write.csv(counts, file = paste0(coverage_dir, 'counts.csv'), quote = FALSE)
write.csv(counts.norm, file = paste0(coverage_dir, 'counts.TPM-normalized.csv'), quote = FALSE)
rm(gene_widths, count_sums)
```

## QC of count table data

```{r, eval=TRUE}
dir.create(image_dir)

dds <- DESeqDataSet(se, design = ~ cultivation_run + genotype + medium)
dds <- DESeq(dds)
rld <- rlog(dds, blind = FALSE)

sampleDists <- dist(t(assay(rld)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- rld$sample_ID
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
pdf(paste0(image_dir, 'heatmap_RNA-seq_sampleDist_rlog.pdf'))
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
dev.off()

pdf(paste0(image_dir, 'PCA_RNA-seq_rlog.pdf'))
plotPCA(rld, intgroup = c('genotype', 'medium', 'cultivation_run'))
plotPCA(rld, intgroup = 'genotype')
plotPCA(rld, intgroup = 'medium')
plotPCA(rld, intgroup = 'cultivation_run')
dev.off()
```

* Heatmap results - samples do not cluster that well. All YES samples form one cluster, YES+AMM are split. This is likely caused by the fact that the differences between all samples from a particular medium are just minor.

* PCA results - no batch effects; medium is the strongest differentiating factor; the response to AMM is more varied in the mga2_cbf11 double mutants (between replicates) and seems to differ from the mga2 strain


## DEG calling

```{r, eval=TRUE}
# Benjamini-Hochberg p value adjustment (FDR)
padj_threshold <- 0.05

dir.create(deg_dir)

# mga2_cbf11-YES vs mga2-YES
dds <- DESeqDataSet(se, design = ~ cultivation_run + genotype)
dds <- dds[, dds$medium =='YES'] # remove YES+ammonium samples ('YES_AMM')
dds$medium <- droplevels(dds$medium) # remove factor levels without samples
dds$genotype <- relevel(dds$genotype, ref = 'mga2') # change order of factor levels to get mga2_cbf11/mga2 fold change (= define the reference)
dds <- DESeq(dds)
res <- results(dds, alpha = padj_threshold)
summary(res[res$padj <= padj_threshold & !is.na(res$padj), ])
res <- cbind(counts[, 'gene_name'], res)
colnames(res)[1] <- 'gene_name'
write.csv(res, file = paste0(deg_dir, 'DESeq2results_mga2_cbf11-YES_vs_mga2-YES.csv'), quote = FALSE)
res.sig <- res[which(res$padj <= padj_threshold), ]
res.sig <- res.sig[order(res.sig$log2FoldChange), ]
write.csv(res.sig, file = paste0(deg_dir, 'DESeq2results_mga2_cbf11-YES_vs_mga2-YES.SIG.csv'), quote = FALSE)

# mga2_cbf11-YES+AMM vs mga2-YES+AMM
dds <- DESeqDataSet(se, design = ~ cultivation_run + genotype)
dds <- dds[, dds$medium =='YES_AMM']
dds$medium <- droplevels(dds$medium)
dds$genotype <- relevel(dds$genotype, ref = 'mga2')
dds <- DESeq(dds)
res <- results(dds, alpha = padj_threshold)
summary(res[res$padj <= padj_threshold & !is.na(res$padj), ])
res <- cbind(counts[, 'gene_name'], res)
colnames(res)[1] <- 'gene_name'
write.csv(res, file = paste0(deg_dir, 'DESeq2results_mga2_cbf11-YES+AMM_vs_mga2-YES+AMM.csv'), quote = FALSE)
res.sig <- res[which(res$padj <= padj_threshold), ]
res.sig <- res.sig[order(res.sig$log2FoldChange), ]
write.csv(res.sig, file = paste0(deg_dir, 'DESeq2results_mga2_cbf11-YES+AMM_vs_mga2-YES+AMM.SIG.csv'), quote = FALSE)

# mga2-YES+AMM vs mga2-YES
dds <- DESeqDataSet(se, design = ~ cultivation_run + medium)
dds <- dds[, dds$genotype =='mga2']
dds$genotype <- droplevels(dds$genotype)
dds$medium <- relevel(dds$medium, ref = 'YES')
dds <- DESeq(dds)
res <- results(dds, alpha = padj_threshold)
summary(res[res$padj <= padj_threshold & !is.na(res$padj), ])
res <- cbind(counts[, 'gene_name'], res)
colnames(res)[1] <- 'gene_name'
write.csv(res, file = paste0(deg_dir, 'DESeq2results_mga2-YES+AMM_vs_mga2-YES.csv'), quote = FALSE)
res.sig <- res[which(res$padj <= padj_threshold), ]
res.sig <- res.sig[order(res.sig$log2FoldChange), ]
write.csv(res.sig, file = paste0(deg_dir, 'DESeq2results_mga2-YES+AMM_vs_mga2-YES.SIG.csv'), quote = FALSE)

# mga2_cbf11-YES+AMM vs mga2_cbf11-YES
dds <- DESeqDataSet(se, design = ~ cultivation_run + medium)
dds <- dds[, dds$genotype =='mga2_cbf11']
dds$genotype <- droplevels(dds$genotype)
dds$medium <- relevel(dds$medium, ref = 'YES')
dds <- DESeq(dds)
res <- results(dds, alpha = padj_threshold)
summary(res[res$padj <= padj_threshold & !is.na(res$padj), ])
res <- cbind(counts[, 'gene_name'], res)
colnames(res)[1] <- 'gene_name'
write.csv(res, file = paste0(deg_dir, 'DESeq2results_mga2_cbf11-YES+AMM_vs_mga2_cbf11-YES.csv'), quote = FALSE)
res.sig <- res[which(res$padj <= padj_threshold), ]
res.sig <- res.sig[order(res.sig$log2FoldChange), ]
write.csv(res.sig, file = paste0(deg_dir, 'DESeq2results_mga2_cbf11-YES+AMM_vs_mga2_cbf11-YES.SIG.csv'), quote = FALSE)
```

***

```{r}
sessionInfo()
```
